\begin{titlepage}
\vspace*{1.00in}
\begin{center}
\Large
{\bf The Emerald Programming Language\footnote{This work was 
supported in part
by the National Science Foundation under Grants No. 
MCS-8004111, DCR-8420945, and CCR-8700106, by
K{\o}benhavns Universitet (the University of Copenhagen), Denmark under
Grant J.nr.\ 574-2,2, by Digital Equipment Corporation
External Research Grants, by an IBM Graduate Fellowship,
and a dissertation award from Microsoft Corporation.}} \\[0.08in]
\large
{\bf REPORT} \\
\large
\vspace{2.5ex}
Norman C. Hutchinson, Rajendra K. Raj, \\
Andrew P. Black, Henry M. Levy, and Eric Jul\footnote{Authors' current 
addresses:  
Norman Hutchinson, Department of Computer Science, University of
Arizona, Tucson, AZ 85721.  
Andrew Black, Digital Equipment Corporation, 550 King Street, Littleton, MA 01460.
Eric Jul, DIKU, Department of Computer Science, University of Copenhagen, 
Universitetsparken 1, DK-2100 Copenhagen, Denmark.} \\
\vspace{2.0ex}
%\vspace{2.3ex}
\normalsize
{\em Department of Computer Science \\
University of Washington \\
Seattle, WA 98195} \\
\vspace{2.0ex}
%\vspace{2.5ex}
Technical Report 87-10-07\\
October 1987\\
(Revised August 1988)
\end{center}
\vspace*{0.4in}
\begin{center}
\parbox{6.0in}{
{\bf Previously also released as:}\\[0.15in]
DIKU Report No. 87/22, Department of Computer Science, University of Copenhagen, Denmark.\\
T.R. 87/29, Department of Computer Science, University of Arizona, Tucson, AZ 85721.\\
}
\end{center}
\vspace*{0.2in}

\begin{center}
\parbox{6.0in}{
{\bf Abstract:}\\[0.15in]
The programming language Emerald was designed and developed 
to demonstrate that the
object-based style of programming can be incorporated both elegantly and
efficiently in the distributed programming environment.
At the same time, Emerald is an '80s programming language providing excellent
features for abstraction and polymorphism. Primarily a language for
distributed environments, Emerald includes features for dealing
with the location of objects,
and extends exception handling mechanisms for recovering from partial failures
of distributed systems. This report presents an overview of
Emerald and a concise description of its language constructs.
}
\end{center}
\vspace*{0.2in}
\mbox{}
\end{titlepage}
\addtocounter{footnote}{-2}
